Future challenges for Research Software Engineering

### 2025-09-19 correspondence/rebuttal-1.md

> I like the idea and the elaboration of the paper, but fail to see relevant contributions in the individual sections, in particular comes to differentiation from Software Engineering. With that respect, related work in Software Engineering already addresses a lot of aspects mentioned in the paper and should be referenced and acknowledged appropriately as it shapes expectations and future work. I'd love to see a better elaboration of the paper, as again it is a relevant and interesting approach.

This is a fair point.
We have familiarized ourselves with the SE literature on this subject and have drawn
parallels between projections made in the SE space and made in the RSE space.
It was particularly interesting to see the overlap between our manuscript
and Katz's 2023 paper ``Research Software Engineering in 2030''.

Many participants of the workshop are scientists by training, and therefore have slightly
different perspectives on the future of RSEng due to the environment they work in.
They also use a different vocabulary compared to software engineers.
The addition of SE literature is definitively an improvement for the manuscript.

> Research Software Engineering and Research Software Engineer ae occasionally written in full when the abbreviation could be used

We have used the abbreviation now in all places except where it is part of a name or a paublication. Also we refrained from starting the intro with an abbreviation.

> Why is Research Software capitalised?

Now it is lower-case, except in names or publication titles.

> The titles in section 2 aren’t consistent in how they are capitalised

Thanks for spotting this, now hte titles are properly capitalised.

### 2025-09-19 future_challenges_for_RSEng.tex
sort out Box-6

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
while the foundational competencies of a research software engineer have just recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how Research Software Engineering should, or could, develop.

@@

\section{Aspects of this future}
\begin{whatis}{}{whatisexplanation}
In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``Goal'' boxes.
\end{whatis}

Looking ahead over a 10-year time frame, we hope that the RSE community will have
contributed to making research software and its outputs
more sustainable, robust and maintainable.
We also hope that Research Software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
We see this happening in a number of ways,
particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
However, we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
Developing software tools and scripted workflows is becoming an increasingly important part of everyday
research activities.
Therefore, it is vital to improve digital skills of all researchers.
This will also allow RSEs to focus on other more advanced and specialist areas.
In this context, we do not foresee the demise of the RSE as an independent role.
However, we do see the work of RSEs developing and shifting as the research community changes~\cite{ChueHong2025}.

@@

\begin{story}{Introduction}{intro}
In order to give a more concrete setting, imagine a world ten years from now.
Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
Over this time frame Kim~\cite{Anzt2021} and Kay~\cite{Goth2024},
our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sun blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems that have been solved by then?
Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
And what are the issues that have newly emerged during this time?
In order to bring this future to life, these two personas will accompany us through the rest of this paper and we will shed light on aspects
of their future in respective small story boxes.
\end{story}

@@

\subsection{Development of the RSE community}
\begin{whatis}{}{development-rse-community}
National RSE communities are increasingly collaborating with local partners,
fostering stronger networks across countries.
Regular international conferences further amplify these connections,
promoting the global exchange of ideas. The introduction of RSE Master's programmes,
coupled with a shift in culture towards valuing Research Software,
has significantly elevated its profile.
This growing recognition has led to a surge in membership of national societies,
reflecting the increasing importance of RSEng in the research landscape.
\end{whatis}
In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSE positions to become more easily accessible,
possibly through centrally-funded long-term positions at research institutions,
such as in the bwRSE4HPC service\footnote{\url{https://www.bwrse4hpc.de/68.php}}.
The community of RSEs will get a significant boost in their acceptance
if RSE Master's programmes become available at various universities.
With increased numbers, the visibility of the RSE community to industry will be enhanced.
\begin{story}{Development of RSE community}{RSEcommunity}
Kay stands proudly at the podium of the first International Research
Software Engineering Conference held in the Southern Hemisphere,
a historic moment for both her career and the global RSE community.
Having moved to South Africa, she speaks passionately about the unique
challenges researchers in the Global South face: limited access
to high-performance computing, unreliable funding,
and the persistent digital divide.
Her keynote highlights how local innovation thrives despite constraints,
often out of necessity rather than choice.
The audience, a diverse mix of global experts, listens intently as she calls
for equitable collaborations, not just inclusion as an afterthought.
\end{story}

\subsection{Impact of AI and automation}
\label{sec:ai}
\begin{whatis}{}{impact-ai-automation}
RSEs understand how AI models work and AI-RSE specialists know how to tune and extend them.
With these skills they can help researchers reap their benefits,
while clearly knowing their limits.
\end{whatis}
Generative AI tools, such as GitHub Copilot~\cite{Friedmann2021},
are being added to software development tools~\cite{Alenezi2025}.
These AI-powered code assistants promise to increase productivity
by automatically generating code, and writing test cases and documentation~\cite{Banh2025}.
Software developers use them to reduce the number of keystrokes,
focus on higher-level design rather than on idiosyncrasies,
and recall syntax without having to consult the documentation~\cite{Liang2024}.
Code assistants have the potential to help novice or even non-expert programmers
write code~\cite{Feldman2024} thus enabling non-experts to participate in research
and software creation, and fostering greater collaboration across disciplines.
However, when these tools are used in practice, many users report increased cognitive
load and frustration stemming from the difficulty to prompt the system effectively
and from debugging the generated code~\cite{Simkute2025}.
Many factors can influence the adoption of AI tools~\cite{Russo2024}.
The potential of AI to be a democratising technology can only be realised
when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
We also note members of the SE community have called for improving LLM literacy
by adapting SE curricula~\cite{Kirova2024}.

@@

\subsection{Ethics and social considerations}
\begin{whatis}{}{ethics}
The principles of Good Scientific Practice~\cite{dfg_gsp} have been a cornerstone in the conduct of research.
But their relevance hinges on constant updates in light of how research is performed.
By 2035, we want to have achieved accepted rules on the use of AI in research tasks and software development.
Open Science and Open Research are by then well-established in academia, and supported by respective Open Research Software
strategies of the political actors. We also hope that Data Awareness and Data Protection are still values
that are protected.
\end{whatis}

The ethics and values of the Research Software Engineer are grounded in their practice.
They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients~\cite{Goth2024}.

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{}{academic}
We expect research software to be a central part of scientific practice, on the same level as data and publications.
This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
Additionally, researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

@@

\section{What is in this future for RSEs?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?
In the previous sections, we have tried to list some of the trends that we foresee.
The material presented here is based on the outputs of a workshop, and extended with the authors' ideas.
Nevertheless, there are some things that we declared out of scope here.
Among that is the ongoing digital transformation in wider society.
Digital competencies in the general population continue to increase,
making the jump to software development easier than before.
But this digitalisation will also lead to changes in the way societies
interact and therefore will translate into changes in how academia perceives itself.
We have also not sufficiently delved into the interaction with industry.
There are also some trends that we believe will stay with us over the coming years.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, arising from technologies such as generally available AI systems.

>>>>

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
while the foundational competencies of an RSE have just recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.

@@

\section{Aspects of this Future}
\begin{whatis}{}{whatisexplanation}
In order to list what has been achieved in a certain aspect within the RSE community of the future of 2035 we use these ``Goal'' boxes.
\end{whatis}

Looking ahead over a 10-year time frame, we hope that the RSE community will have
contributed to making research software and its outputs
more sustainable, robust and maintainable.
We also hope that research software becomes an accepted research artefact on an equal footing with data publications and traditional text publications.
We see this happening in a number of ways,
particularly through movements such as DORA~\cite{DORA} and CoARA~\cite{COARA} which are advocating for a change in the way that research is assessed and outputs recognised.
However, we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
Developing software tools and scripted workflows is becoming an increasingly important part of everyday
research activities.
Therefore, it is vital to improve digital skills of all researchers.
This will also allow RSEs to focus on other more advanced and specialist areas.
In this context, we do not foresee the demise of the RSE as an independent role.
However, we do see the work of RSEs developing and shifting as the research community changes~\cite{ChueHong2025}.

@@

\begin{story}{Introduction}{intro}
In order to give a more concrete setting, imagine a world ten years from now.
Not too overwhelmingly far ahead in the future, but distant enough for some changes to have occurred.
Over this time frame Kim~\cite{Anzt2021} and Kay~\cite{Goth2024},
our fictional RSEs who we have used to illustrate specific scenarios in our earlier work,
were able to ride the waves of their career in RSEng and are now pondering early retirement on a sun blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems that have been solved by then?
Which tasks are still relevant, where they can be happy that a younger generation is now taking up the baton and carrying forward their work?
And what are the issues that have newly emerged during this time?
In order to bring this future to life, these two personas will accompany us through the rest of this paper and we will shed light on aspects
of their future in respective small story boxes.
\end{story}

@@

\subsection{Development of the RSE Community}
\begin{whatis}{}{development-rse-community}
National RSE communities are increasingly collaborating with local partners,
fostering stronger networks across countries.
Regular international conferences further amplify these connections,
promoting the global exchange of ideas. The introduction of RSE Master's programmes,
coupled with a shift in culture towards valuing research software,
has significantly elevated its profile.
This growing recognition has led to a surge in membership of national societies,
reflecting the increasing importance of RSEng in the research landscape.
\end{whatis}
In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSE positions to become more easily accessible,
possibly through centrally-funded long-term positions at research institutions,
such as in the bwRSE4HPC service\footnote{\url{https://www.bwrse4hpc.de/68.php}}.
The community of RSEs will get a significant boost in their acceptance
if RSE Master's programmes become available at various universities.
With increased numbers, the visibility of the RSE community to industry will be enhanced.
\begin{story}{Development of RSE community}{RSEcommunity}
Kay stands proudly at the podium of the first International Research
Software Engineering Conference held in the Southern Hemisphere,
a historic moment for both her career and the global RSE community.
Having moved to South Africa, she speaks passionately about the unique
challenges researchers in the Global South face: limited access
to high-performance computing, unreliable funding,
and the persistent digital divide.
Her keynote highlights how local innovation thrives despite constraints,
often out of necessity rather than choice.
The audience, a diverse mix of global experts, listens intently as she calls
for equitable collaborations, not just inclusion as an afterthought.
\end{story}

\subsection{Impact of AI and Automation}
\label{sec:ai}
\begin{whatis}{}{impact-ai-automation}
RSEs understand how AI models work and AI-RSE specialists know how to tune and extend them.
With these skills they can help researchers reap their benefits,
while clearly knowing their limits.
\end{whatis}
Generative AI tools, such as GitHub Copilot~\cite{Friedmann2021},
are being added to software development tools~\cite{Alenezi2025}.
These AI-powered code assistants promise to increase productivity
by automatically generating code, and writing test cases and documentation~\cite{Banh2025}.
Software developers use them to reduce the number of keystrokes,
focus on higher-level design rather than on idiosyncrasies,
and recall syntax without having to consult the documentation~\cite{Liang2024}.
Code assistants have the potential to help novice or even non-expert programmers
write code~\cite{Feldman2024} thus enabling non-experts to participate in research
and software creation, and fostering greater collaboration across disciplines.
However, when these tools are used in practice, many users report increased cognitive
load and frustration stemming from the difficulty to prompt the system effectively
and from debugging the generated code~\cite{Simkute2025}.
Many factors can influence the adoption of AI tools~\cite{Russo2024}.
The potential of AI to be a democratising technology can only be realised
when the algorithmic divide along with the digital divide is addressed~\cite{Yu2020}.
We also note members of the SE community have called for improving LLM literacy
by adapting SE curricula~\cite{Kirova2024}.

@@

\subsection{Ethics and Social Considerations}
\begin{whatis}{}{ethics}
The principles of Good Scientific Practice~\cite{dfg_gsp} have been a cornerstone in the conduct of research.
But their relevance hinges on constant updates in light of how research is performed.
By 2035, we want to have achieved accepted rules on the use of AI in research tasks and software development.
Open Science and Open Research are by then well-established in academia, and supported by respective open research software
strategies of the political actors. We also hope that Data Awareness and Data Protection are still values
that are protected.
\end{whatis}

The ethics and values of the RSE are grounded in their practice.
They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients~\cite{Goth2024}.

@@

\subsection{How have Academic Institutions Transformed by Then?}
\begin{whatis}{}{academic}
We expect research software to be a central part of scientific practice, on the same level as data and publications.
This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
Additionally, researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

@@

\section{What is in this Future for RSEs?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?
In the previous sections, we have tried to list some of the trends that we foresee.
The material presented here is based on the outputs of a workshop, and extended with the authors' ideas.
Nevertheless, there are some things that we declared out of scope here.
Among that is the ongoing digital transformation in wider society.
Digital competencies in the general population continue to increase,
making the jump to software development easier than before.
But this digitalisation will also lead to changes in the way societies
interact and therefore will translate into changes in how academia perceives itself.
We have also not sufficiently delved into the interaction with industry.
There are also some trends that we believe will stay with us over the coming years.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, arising from technologies such as generally available AI systems.
